A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-09-28 Manuscript/Figs/SI.Rmd
MEE version

---
title:

author:

@@

csl: pnas.csl
bibliography: Lambda-Refs.bib
---

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("SI-Title.png")  
```
\hfill\break
\hfill\break

## Supplementary Information for 
\hfill\break

A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal \hfill\break
\hfill\break
Dean C. Adams, Erica K. Baken, and Michael L. Collyer \hfill\break

\hfill\break
Dean C. Adams\hfill\break
Email:  dcadams@iastate.edu\hfill\break

## This PDF file includes:
\hfill\break
- Supplementary text 
- Figures S1 to S20 
- SI References

@@

Here we examine the statistical properties of tests for comparing the relative strength of phylogenetic signal, using $\hat{Z}_{12}$. Simulations were performed using pure-birth trees of different sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). For each combination of $n$ and $\lambda$, 20 pure-birth trees were generated and 50 traits were simulated on each phylogeny under a Brownian motion model of evolution. Next, the phylogenetic signal between traits was compared in pairwise fashion for all $50*49/2 = 1225$ combinations, using $\hat{Z}_{12}$. The proportion of significant results provided an estimate of the type I error or model misspecification. Specifically, type I error was evaluated when traits were simulated using no input phylogenetic signal (i.e., $\lambda = 0$). Likewise, model misspecification was evaluated for data simulated with some known, but equal, level of phylogenetic signal (i.e., $\lambda > 0$). \hfill\break

>>>>

---
title: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal
author: Dean C. Adams, Erica K. Baken, and Michael L. Collyer

@@

csl: methods-in-ecology-and-evolution.csl
bibliography: Lambda-Refs.bib
---

@@

# Additional Simulation Results

@@

Here we examine the statistical properties of tests for comparing the relative strength of phylogenetic signal, using $\hat{Z}_{12}$. Simulations were performed using pure-birth trees of different sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). For each combination of $n$ and $\lambda$, 2 pure-birth trees were generated and 50 traits were simulated on each phylogeny under a Brownian motion model of evolution. Next, the phylogenetic signal between traits was compared in pairwise fashion for all $50*49/2 = 1225$ combinations, using $\hat{Z}_{12}$. The proportion of significant results provided an estimate of the type I error or model misspecification. Specifically, type I error was evaluated when traits were simulated using no input phylogenetic signal (i.e., $\lambda = 0$). Likewise, model misspecification was evaluated for data simulated with some known, but equal, level of phylogenetic signal (i.e., $\lambda > 0$). \hfill\break

### 2020-09-28 Manuscript/Figs/SI.tex
MEE version

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\includegraphics[width=0.95\linewidth]{SI-Title} \hfill\break
\hfill\break

\hypertarget{supplementary-information-for}{%
\subsection{Supplementary Information
for}\label{supplementary-information-for}}

\hfill\break

A Standardized Effect Size for Evaluating and Comparing the Strength of
Phylogenetic Signal \hfill\break \hfill\break Dean C. Adams, Erica K.
Baken, and Michael L. Collyer \hfill\break

\hfill\break

Dean C. Adams\hfill\break Email:
\href{mailto:dcadams@iastate.edu}{\nolinkurl{dcadams@iastate.edu}}\hfill\break

\hypertarget{this-pdf-file-includes}{%
\subsection{This PDF file includes:}\label{this-pdf-file-includes}}

\hfill\break

\begin{itemize}
\tightlist
\item
  Supplementary text
\item
  Figures S1 to S20
\item
  SI References
\end{itemize}

@@

We also investigaged the effect of unresolved phylogenies on estimates
of phylogenetic signal by adjusting pure-birth trees to have 20\%
collapsed nodes, following the procedures of (1). Results from
simulations on phylogenies containing polytomies mirrored those found on
fully resolved trees. For \(\hat{\lambda}\), the mean value increased
with increasing input signal, but was negatively biased, and was less
than the input value across most of its range (Fig. S8: black line).
Additionally, the precision of \(\hat{\lambda}\) varied with differing
input levels, with the greatest variation found at intermediate values
of \(\lambda\) (Fig. S8 red line). Finally, the distribution of
\(\hat{\lambda}\) was not normal, and became more skewed at more extreme
values of \(\lambda\) (Fig. S8 blue line). \hfill\break

@@

With respect to model performance, input parameters (\(\beta\)) were
well estimated in phylogenetic regression and ANOVA when \(\lambda\) was
incorporated (Figs S17 \& S18), implying that \(\lambda\) may be used as
a tuning parameter for these models. Additionally, type I error and
statistical power were unaffected by the inclusion of \(\lambda\) (Figs
S19 \& S20), though power was reduced at smaller sample sizes in
phylogenetic ANOVA, mirroring prior findings of (2). These results
generally confirmed earlier findings of (3) for phylogenetic regression,
and extend them to the case of phylogenetic ANOVA.

@@

Here we examine the statistical properties of tests for comparing the
relative strength of phylogenetic signal, using \(\hat{Z}_{12}\).
Simulations were performed using pure-birth trees of different sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylogenetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). For each
combination of \(n\) and \(\lambda\), 20 pure-birth trees were generated
and 50 traits were simulated on each phylogeny under a Brownian motion
model of evolution. Next, the phylogenetic signal between traits was
compared in pairwise fashion for all \(50*49/2 = 1225\) combinations,
using \(\hat{Z}_{12}\). The proportion of significant results provided
an estimate of the type I error or model misspecification. Specifically,
type I error was evaluated when traits were simulated using no input
phylogenetic signal (i.e., \(\lambda = 0\)). Likewise, model
misspecification was evaluated for data simulated with some known, but
equal, level of phylogenetic signal (i.e., \(\lambda > 0\)).
\hfill\break

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-MolinaVenegas2017}{}%
1. Molina-Venegas R, Rodríguez MA (2017) Revisiting phylogenetic signal;
strong or negligible impacts of polytomies and branch length
information? \emph{BMC evolutionary biology} 17(1):53.

\leavevmode\hypertarget{ref-AdamsCollyer2018b}{}%
2. Adams DC, Collyer ML (2018) Phylogenetic anova: Group-clade
aggregation, biological challenges, and a refined permutation procedure.
\emph{Evolution} 72(6):1204--1215.

\leavevmode\hypertarget{ref-Revell2010}{}%
3. Revell LJ (2010) Phylogenetic signal and linear regression on species
data. \emph{Methods in Ecology and Evolution} 1:319--329.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal},
  pdfauthor={Dean C. Adams, Erica K. Baken, and Michael L. Collyer},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{A Standardized Effect Size for Evaluating and Comparing the Strength of
Phylogenetic Signal}
\author{Dean C. Adams, Erica K. Baken, and Michael L. Collyer}
\date{}

\begin{document}
\maketitle

\hypertarget{additional-simulation-results}{%
\section{Additional Simulation
Results}\label{additional-simulation-results}}

@@

We also investigaged the effect of unresolved phylogenies on estimates
of phylogenetic signal by adjusting pure-birth trees to have 20\%
collapsed nodes, following the procedures of (Molina-Venegas \&
Rodríguez, 2017). Results from simulations on phylogenies containing
polytomies mirrored those found on fully resolved trees. For
\(\hat{\lambda}\), the mean value increased with increasing input
signal, but was negatively biased, and was less than the input value
across most of its range (Fig. S8: black line). Additionally, the
precision of \(\hat{\lambda}\) varied with differing input levels, with
the greatest variation found at intermediate values of \(\lambda\) (Fig.
S8 red line). Finally, the distribution of \(\hat{\lambda}\) was not
normal, and became more skewed at more extreme values of \(\lambda\)
(Fig. S8 blue line). \hfill\break

@@

With respect to model performance, input parameters (\(\beta\)) were
well estimated in phylogenetic regression and ANOVA when \(\lambda\) was
incorporated (Figs S17 \& S18), implying that \(\lambda\) may be used as
a tuning parameter for these models. Additionally, type I error and
statistical power were unaffected by the inclusion of \(\lambda\) (Figs
S19 \& S20), though power was reduced at smaller sample sizes in
phylogenetic ANOVA, mirroring prior findings of (Adams \& Collyer,
2018). These results generally confirmed earlier findings of (Revell,
2010) for phylogenetic regression, and extend them to the case of
phylogenetic ANOVA.

@@

Here we examine the statistical properties of tests for comparing the
relative strength of phylogenetic signal, using \(\hat{Z}_{12}\).
Simulations were performed using pure-birth trees of different sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylogenetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). For each
combination of \(n\) and \(\lambda\), 2 pure-birth trees were generated
and 50 traits were simulated on each phylogeny under a Brownian motion
model of evolution. Next, the phylogenetic signal between traits was
compared in pairwise fashion for all \(50*49/2 = 1225\) combinations,
using \(\hat{Z}_{12}\). The proportion of significant results provided
an estimate of the type I error or model misspecification. Specifically,
type I error was evaluated when traits were simulated using no input
phylogenetic signal (i.e., \(\lambda = 0\)). Likewise, model
misspecification was evaluated for data simulated with some known, but
equal, level of phylogenetic signal (i.e., \(\lambda > 0\)).
\hfill\break

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-AdamsCollyer2018b}{}%
Adams, D. C., \& Collyer, M. L. (2018). Phylogenetic anova: Group-clade
aggregation, biological challenges, and a refined permutation procedure.
\emph{Evolution}, \emph{72}(6), 1204--1215.

\leavevmode\hypertarget{ref-MolinaVenegas2017}{}%
Molina-Venegas, R., \& Rodríguez, M. A. (2017). Revisiting phylogenetic
signal; strong or negligible impacts of polytomies and branch length
information? \emph{BMC Evolutionary Biology}, \emph{17}(1), 53.

\leavevmode\hypertarget{ref-Revell2010}{}%
Revell, L. J. (2010). Phylogenetic signal and linear regression on
species data. \emph{Methods in Ecology and Evolution}, \emph{1},
319--329. Journal Article.

### 2020-09-28 Manuscript/OldVsns/PNAS-Version/PNAS-Adams-Submitted.tex
MEE version

To demonstrate the utility of \(\hat{Z}_{12}\), we compared \(Z_K\) for
two ecologically-relevant traits in plethodontid salamander (Fig. 4):
surface area to volume ratios (SA:V) and relative body width
(\(\frac{BW}{SVL}\)) (41, 42). While both traits contained significant
phylogenetic signal, tests based on \(\hat{Z}_{12}\) revealed that the
degree of phylogenetic signal was significantly stronger in SA:V
(\(\hat{Z}_{12}=4.13\); \(P=0.000036\): Fig. 4). Biologically, this
observation may be interpreted by the fact that the tropical species --
which form a monophyletic group within plethodontids -- display greater
variation in SA:V, which covaries with disparity in their climatic
niches (42). Thus, greater phylogenetic signal in SA:V is to be
expected.

>>>>

To demonstrate the utility of \(\hat{Z}_{12}\), we compared \(Z_K\) for
two ecologically-relevant traits in plethodontid salamander (Fig. 4):
surface area to volume ratios (SA:V) and relative body width
(\(\frac{BW}{SVL}\)) (41, 42). While both traits contained significant
phylogenetic signal, tests based on \(\hat{Z}_{12}\) revealed that the
degree of phylogenetic signal was significantly stronger in SA:V
(\(\hat{Z}_{12}=16.51\); \(P< 0.00001\): Fig. 4). Biologically, this
observation may be interpreted by the fact that the tropical species --
which form a monophyletic group within plethodontids -- display greater
variation in SA:V, which covaries with disparity in their climatic
niches (42). Thus, greater phylogenetic signal in SA:V is to be
expected.

### 2020-09-28 Manuscript/SI.Rmd
MEE version

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S1.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S2.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S3.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S4.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S5.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S6.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S7.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S8.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S9.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S10.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S11.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S12.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S13.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S14.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S15.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S16.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S17.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S18.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S19.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("fig.S20.png")  
```

@@

```{r, echo = FALSE, out.height="50%", align= "center"}
include_graphics("fig.S21.png")  
```

\textbf{Figure S21}. Type I error rates and model misspecification for $\hat{Z}_{12}$. Type I error is found on the far-left of the plot ($\lambda = 0$). The solid black line represents the average type I error and false discovery rate across all levels of input signal.

>>>>

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S1.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S2.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S3.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S4.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S5.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S6.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S7.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S8.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S9.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S10.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S11.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S12.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S13.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S14.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S15.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S16.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S17.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S18.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S19.png")  
```

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("Figs/fig.S20.png")  
```

@@

```{r, echo = FALSE, out.height="50%", align= "center"}
include_graphics("Figs/fig.S21.png")  
```

\textbf{Figure S21}. Type I error rates and model misspecification for $\hat{Z}_{12}$. Type I error is found on the far-left of the plot ($\lambda = 0$).
